# Technical reference {#reference}

This chapter is the formal reference for the schedule file format and the two network interfaces a device exposes. Chapter \@ref(manual) covers the same ground in practical terms; if you are running an experiment rather than writing software against the platform, start there.

## Schedule file format {#schedule-format}

A schedule is a JSON document with a single `scheduleConfig` object, which carries the run-wide configuration and a `schedule` array of setpoints.

### `scheduleConfig` fields {-}

```{r, echo=FALSE}
df <- tibble::tribble(
  ~Field, ~Required, ~Value,
  "`deviceID`", "yes", "The device's MAC suffix (case-sensitive), or `\"all\"` (case-insensitive) to allow any device. A schedule whose `deviceID` matches neither is **rejected on load**. See Section \\@ref(device-id).",
  "`scheduleName`", "yes", "Free-text label identifying the schedule.",
  "`interpolation`", "yes", "`\"linear\"`, `\"step\"` (instant changes) or `\"cubic\"` (Hermite spline).",
  "`timeType`", "yes", "`\"relative\"` (time since device start) or `\"absolute\"` (**UTC**, formatted `YYYY_MM_DD HH_mm_SS`).",
  "`timeUnit`", "yes", "`\"hours\"`, `\"minutes\"` or `\"seconds\"`. Ignored for absolute time.",
  "`startTime`", "relative only", "When the schedule starts, in `timeUnit`.",
  "`wifiSSID`", "yes", "Network the device joins at boot.",
  "`wifiPassword`", "yes", "Password for that network.",
  "`imageIntervalMin`", "yes", "Minutes between captures; must be greater than 0.",
  "`loopTime`", "relative only", "Length of one cycle, in `timeUnit`.",
  "`totalLoops`", "yes", "Number of repeats; `-1` repeats forever. For absolute schedules, any value other than 1 causes the date part to be ignored and the schedule treated as a 24-hour loop.",
  "`focusMode`", "no", "`\"Auto\"` (default) or `\"Manual\"`. Case-insensitive.",
  "`focusValue`", "if manual", "Fixed focus position, integer 0 (far) to 255 (close).",
  "`mqttBrokerUri`", "no", "Enables telemetry, e.g. `mqtt://192.168.1.50:1883`.",
  "`mqttUsername`", "no", "Broker username.",
  "`mqttPassword`", "no", "Broker password.",
  "`mqttTopicPrefix`", "no", "Defaults to `dwraph/v1/devices` — see Section \\@ref(mqtt)."
)
knitr::kable(df)
```

### Setpoint fields {-}

```{r, echo=FALSE}
df <- tibble::tribble(
  ~Field, ~Value,
  "`name`", "Label for the setpoint. Not interpreted by the firmware.",
  "`description`", "Free-text description. Not interpreted by the firmware.",
  "`time`", "When this setpoint applies, in `timeUnit`, or an absolute timestamp.",
  "`temperature`", "Target in °C. **Clamped by the firmware to 10--50 °C.**",
  "`light`", "Object with `red`, `blue` and `white`, each 0.0--1.0."
)
knitr::kable(df)
```

The file belongs at `/sdcard/<DEVICE_ID>/schedule.json` on the device's SD card.

## Device identity {#identity}

A device's ID is derived at boot from its factory MAC address: the last six hexadecimal digits, upper case. It is not configurable, and the `deviceID` field of a schedule is a match condition rather than an assignment.

The ID propagates to:

```{r, echo=FALSE}
df <- tibble::tribble(
  ~Where, ~Form,
  "SD card", "`/sdcard/<DEVICE_ID>/`",
  "mDNS hostname", "`dwarph-<DEVICE_ID>`",
  "mDNS service instance", "`DWARPH <DEVICE_ID>`",
  "MQTT topic root", "`<prefix>/<DEVICE_ID>`",
  "REST", "`GET /api/id`"
)
knitr::kable(df)
```

## REST API {#rest-api}

Every device serves an HTTP API. The GUI is built entirely on it, so anything the GUI can do can also be scripted.

```{r, echo=FALSE}
df <- tibble::tribble(
  ~Group, ~Endpoints,
  "Identity and health", "`/api/hello`, `/api/id`, `/api/metadata`, `/api/error`, `/api/restart`",
  "Scheduler", "`/api/schedule/start`, `/api/schedule/stop`, `/api/schedule/reset`",
  "Sensors and actuators", "`/api/temps`, `/api/light/state`, `/api/oled/state`",
  "Imaging", "`/api/image`, `/api/image/latest`, `/api/stream`",
  "SD card", "`/api/sdcard/files`, `/api/sdcard/index`, `/api/sdcard/space`, `/api/sdcard/upload`, `/api/sdcard/delete`",
  "Maintenance", "`/api/firmware/upload`, `/api/mqtt`"
)
knitr::kable(df)
```

`GET /api/mqtt` is the discovery hook: it returns the broker address and topic strings for that device, so a client can subscribe without hard-coding the topic layout.

`/api/firmware/upload` accepts a firmware image over the network, so devices can be updated without being taken apart.

<!-- TODO: document request and response payloads for each endpoint. -->

## MQTT telemetry {#mqtt}

MQTT is optional. Without a broker configured, a device is still fully controllable over REST; what you lose is the once-per-second live update that drives the dashboard cards.

It is enabled per device through the optional `mqtt*` fields in `scheduleConfig` (see Section \@ref(schedule-format)). Each device then publishes on:

```{r, echo=FALSE}
df <- tibble::tribble(
  ~Topic, ~Retained, ~Contents,
  "`<prefix>/<deviceID>/fast`", "no", "State published every second: temperature, setpoint, heater output, light channels, network status, SD usage and scheduler state",
  "`<prefix>/<deviceID>/status`", "yes", "`online` / `offline`",
  "`<prefix>/<deviceID>/meta`", "yes", "Topic metadata"
)
knitr::kable(df)
```

The MQTT client ID is `dwraph-<deviceID>`.

Note that the default topic prefix is spelled **`dwraph`**, not `dwarph`. This is a typo in the current firmware, and the GUI matches it, so the two interoperate correctly --- but it will trip you up when writing your own client or grepping for topics. Set `mqttTopicPrefix` explicitly if you would rather not inherit it.

<!-- TODO: fix the `dwraph` prefix in firmware and GUI together, with a migration note here. -->

## Discovery {#discovery}

Devices announce themselves over mDNS/ZeroConf. The GUI scans for services whose name contains `dwarph` (the filter keyword is configurable) and tracks them as they come and go, so devices appear and disappear from the grid as they are powered on and off.

This is why the GUI has no "add device by address" field, and why nothing is discovered when the laptop and the devices are on different subnets, or on a network that blocks multicast.
